\section{Abstract convergence}
\label{sect.abstract.convergence}

Strong eventual consistency requires \emph{convergence} of all copies of shared state stored on nodes in the distributed system.
It guarantees that whenever two nodes have received the same set of updates to shared state---possibly in a different order---their view of the shared state is always identical.
This definition constrains the values that read operations may return at any time, making it a stronger property than eventual consistency.
By accessing only their local copy of the shared state, nodes can execute read and write operations without waiting for network communication.
Nodes exchange updates asynchronously when a network connection is available.  

We now use Isabelle to formalise the notion of strong eventual consistency.
In this section we do not make any assumptions about networks, data structures, or conflict resolution algorithms; instead, we use an abstract model of operations that may be reordered, and we reason about the properties that those operations must satisfy.
We then provide concrete implementations of that abstract model in later sections.

\subsection{The happens-before relation and causality}\label{sect.happens.before}

The simplest way of achieving convergence is to require all operations to be commutative, but this definition is too strong to be useful for many datatypes.
For example, in a list data structure, an element may first be added and then subsequently removed again.
Although it is possible to make such additions and removals unconditionally commutative, doing so yields counterintuitive semantics \cite{Bieniusa:2012wu,Bieniusa:2012gt}.
Instead, a better approach is to require only \emph{concurrent} operations to commute with each other.
Two operations are concurrent if neither ``knew about'' the other at the time when they were generated.
If one operation happened before another---for example, if the removal of an element from a set knew about the prior addition of that element from the set---then it is reasonable to assume that all nodes will apply the operations in that order (first the addition, then the removal).

The \emph{happens-before} relation, as introduced by \citet{Lamport:1978jq}, captures such causal dependencies between operations.
It can be defined in terms of sending and receiving messages on a network, and we give such a definition in Section~\ref{sect.network}.
However, for now, we keep it abstract, writing $\isa{x} \prec \isa{y}$ to indicate that operation $\isa{x}$ happened before $\isa{y}$, where $\prec$ is a predicate of type $\isacharprime\isa{oper} \mathbin{\isasymRightarrow} \isacharprime\isa{oper} \mathbin{\isasymRightarrow} \isa{bool}$.
In words, $\prec$ can be applied to two operations of some abstract type $\isacharprime\isa{oper}$, returning either $\isa{True}$ or $\isa{False}$.%
\footnote{Note that in the distributed systems literature it is conventional to write the happens-before relation as $\isa{x} \rightarrow \isa{y}$, but we reserve the arrow operator to denote logical implication.}
Our only restriction on the happens-before relation $\prec$ is that it must be a \emph{strict partial order}, that is, it must be irreflexive and transitive, which implies that it is also antisymmetric.
We say that two operations $x$ and $y$ are \emph{concurrent}, written $x \mathbin{\isasymparallel} y$, whenever one does not happen before the other:
$\neg (\isa{x} \prec \isa{y})$ and $\neg (\isa{y} \prec \isa{x})$.
Thus, given any two operations $\isa{x}$ and $\isa{y}$, there are three mutually exclusive ways in which they can be related: either $\isa{x} \prec \isa{y}$, or $\isa{y} \prec \isa{x}$, or $\isa{x} \mathbin{\isasymparallel} \isa{y}$.

As discussed above, the purpose of the happens-before relation is to require that some operations must be applied in a particular order, while allowing concurrent operations to be reordered with respect to each other.
We assume that each node applies operations in some sequential order (a standard assumption for distributed algorithms), and so we can model the execution history of a node as a list of operations.
We can then inductively define a list of operations as being \emph{consistent with the happens-before relation}, or simply \emph{hb-consistent}, as follows:
\vspace{0.35em}
\begin{isabellebody}
\ \ \ \ \ \ \ \ \isacommand{inductive} hb{\isacharunderscore}consistent\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharprime}oper\ list\ {\isasymRightarrow}\ bool{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}hb{\isacharunderscore}consistent\ {\isacharbrackleft}{\isacharbrackright}{\isachardoublequoteclose}\ {\isacharbar}\isanewline
\ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}{\isasymlbrakk}\ hb{\isacharunderscore}consistent\ xs{\isacharsemicolon}\ {\isasymforall}x\ {\isasymin}\ set\ xs{\isachardot}\ {\isasymnot}\ y\ {\isasymprec}\ x\ {\isasymrbrakk}\ {\isasymLongrightarrow}\ hb{\isacharunderscore}consistent\ {\isacharparenleft}xs\ {\isacharat}\ {\isacharbrackleft}y{\isacharbrackright}{\isacharparenright}{\isachardoublequoteclose}
\end{isabellebody}
\vspace{0.35em}
In words: the empty list is hb-consistent; furthermore, given an hb-consistent list $\isa{xs}$, we can append an operation $\isa{y}$ to the end of the list to obtain another hb-consistent list, provided that $\isa{y}$ does not happen-before any existing operation $\isa{x}$ in $\isa{xs}$. As a result, whenever two operations $\isa{x}$ and $\isa{y}$ appear in a hb-consistent list, and $\isa{x}\prec\isa{y}$, then $\isa{x}$ must appear before $\isa{y}$ in the list. However, if $\isa{x}\mathbin{\isasymparallel}\isa{y}$, the operations can appear in the list in either order.

\subsection{Interpretation of operations}\label{sect.ops.interpretation}

We describe the state of a node using an abstract type variable $\isacharprime\isa{state}$.
To model state changes, we assume the existence of an \emph{interpretation} function of type $\isa{interp} \mathbin{\isacharcolon\isacharcolon} \isacharprime\isa{oper} \mathbin{\isasymRightarrow} \isacharprime\isa{state} \mathbin{\isasymRightarrow} \isacharprime\isa{state}\ \isa{option}$, which lifts an operation into a \emph{state transformer}---a function that either maps an old state to a new state, or fails by returning $\isa{None}$.
If $\isa{x}$ is an operation, we also write $\langle\isa{x}\rangle$ for the state transformer obtained by applying $\isa{x}$ to the interpretation function.

Concretely, these definitions are captured in Isabelle with the following locale declaration:
\vspace{0.35em}
\begin{isabellebody}
\ \ \ \ \ \ \ \ \isacommand{locale} happens{\isacharunderscore}before\ {\isacharequal}\ preorder\ hb{\isacharunderscore}weak\ hb\isanewline
\ \ \ \ \ \ \ \ \ \ \isakeyword{for}\ hb{\isacharunderscore}weak\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharprime}oper\ {\isasymRightarrow}\ {\isacharprime}oper\ {\isasymRightarrow}\ bool{\isachardoublequoteclose}\ \ {\isacharparenleft}\isakeyword{infix}\ {\isachardoublequoteopen}{\isasympreceq}{\isachardoublequoteclose}\ {\isadigit{5}}{\isadigit{0}}{\isacharparenright}\ \isakeyword{and}\ hb\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharprime}oper\ {\isasymRightarrow}\ {\isacharprime}oper\ {\isasymRightarrow}\ bool{\isachardoublequoteclose}\ \ {\isacharparenleft}\isakeyword{infix}\ {\isachardoublequoteopen}{\isasymprec}{\isachardoublequoteclose}\ {\isadigit{5}}{\isadigit{0}}{\isacharparenright}\ {\isacharplus}\isanewline
\ \ \ \ \ \ \ \ \ \ \isakeyword{fixes}\ interp\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharprime}oper\ {\isasymRightarrow}\ {\isacharprime}state\ {\isasymRightarrow}\ {\isacharprime}state\ option{\isachardoublequoteclose}\ {\isacharparenleft}{\isachardoublequoteopen}{\isasymlangle}{\isacharunderscore}{\isasymrangle}{\isachardoublequoteclose}\ {\isacharbrackleft}{\isadigit{0}}{\isacharbrackright}\ {\isadigit{1}}{\isadigit{0}}{\isadigit{0}}{\isadigit{0}}{\isacharparenright}
\end{isabellebody}
\vspace{0.35em}
The $\isa{happens-before}$ locale extends the $\isa{preorder}$ locale, which is part of Isabelle's standard library and includes various useful lemmas.
It fixes two constants: a preorder that we call $\isa{hb-weak}$ or $\preceq$, and a strict partial order that we call $\isa{hb}$ or $\prec$.
We are only interested in the strict partial order, so we define $\isa{hb-weak}$ as $\isa{x}\preceq\isa{y} \equiv \isa{x}\prec\isa{y} \vee \isa{x}=\isa{y}$, and otherwise ignore it.

Moreover, the locale fixes the interpretation function $\isa{interp}$ as described above, which means that we assume the existence of a function with the given type signature without specifying an implementation.
The code in parentheses defines the mixfix operators $\preceq$, $\prec$, and $\langle\isa{x}\rangle$ for notational convenience.

Given two operations $\isa{x}$ and $\isa{y}$, we can now define the composition of state transformers: we write $\langle\isa{x}\rangle \mathbin{\isasymrhd} \langle\isa{y}\rangle$ to denote the state transformer that first applies the effect of $\isa{x}$ to some state, and then applies the effect of $\isa{y}$ to the result.
If either $\langle\isa{x}\rangle$ or $\langle\isa{y}\rangle$ fails, the combined state transformer also fails.
The operator $\isasymrhd$ is a specialised form of the \emph{Kleisli arrow composition}, which we define as:
\vspace{0.35em}
\begin{isabellebody}
\ \ \ \ \ \ \ \ \isacommand{definition} kleisli\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharparenleft}{\isacharprime}a\ {\isasymRightarrow}\ {\isacharprime}a\ option{\isacharparenright}\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}a\ {\isasymRightarrow}\ {\isacharprime}a\ option{\isacharparenright}\ {\isasymRightarrow}\ {\isacharparenleft}{\isacharprime}a\ {\isasymRightarrow}\ {\isacharprime}a\ option{\isacharparenright}{\isachardoublequoteclose}\ {\isacharparenleft}\isakeyword{infixr}\ {\isachardoublequoteopen}{\isasymrhd}{\isachardoublequoteclose}\ {\isadigit{6}}{\isadigit{5}}{\isacharparenright}\ \isakeyword{where}\isanewline
\ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}f\ {\isasymrhd}\ g\ {\isasymequiv}\ {\isasymlambda}x{\isachardot}\ f\ x\ {\isasymbind}\ {\isacharparenleft}{\isasymlambda}y{\isachardot}\ g\ y{\isacharparenright}{\isachardoublequoteclose}
\end{isabellebody}
\vspace{0.35em}
Here, $\isasymbind$ is the \emph{monadic bind} operation, defined on the option type that we are using to implement partial functions.
We can now define a function $\isa{apply-operations}$ that composes an arbitrary list of operations into a state transformer.
We first map $\isa{interp}$ across the list to obtain a state transformer for each operation, and then collectively compose them using the Kleisli arrow composition combinator:
\vspace{0.35em}
\begin{isabellebody}
\ \ \ \ \ \ \ \ \isacommand{definition} apply{\isacharunderscore}operations\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharprime}oper\ list\ {\isasymRightarrow}\ {\isacharprime}state\ {\isasymRightarrow}\ {\isacharprime}state\ option{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}apply{\isacharunderscore}operations\ es\ {\isasymequiv}\ foldl\ {\isacharparenleft}op\ {\isasymrhd}{\isacharparenright}\ Some\ {\isacharparenleft}map\ interp\ es{\isacharparenright}{\isachardoublequoteclose}
\end{isabellebody}
\vspace{0.35em}
The result is a state transformer that applies the interpretation of each of the operations in the list, in left-to-right order, to some initial state.
If any of the operations fails, the entire composition returns $\isa{None}$.

\subsection{Commutativity and convergence}\label{sect.ops.commute}

We say that two operations $\isa{x}$ and $\isa{y}$ \emph{commute} whenever $\langle\isa{x}\rangle \mathbin{\isasymrhd} \langle\isa{y}\rangle = \langle\isa{y}\rangle \mathbin{\isasymrhd} \langle\isa{x}\rangle$, i.e. when we can swap the order of the composition of their interpretations without changing the resulting state transformer.
For our purposes, requiring that this property holds for \emph{all} pairs of operations is too strong.
Rather, the commutation property is only required to hold for operations that are concurrent, as captured in the next definition:
\vspace{0.35em}
\begin{isabellebody}
\ \ \ \ \ \ \ \ \isacommand{definition} concurrent{\isacharunderscore}ops{\isacharunderscore}commute\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharprime}oper\ list\ {\isasymRightarrow}\ bool{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ \ \ \ \ \ \ \ \ {\isachardoublequoteopen}concurrent{\isacharunderscore}ops{\isacharunderscore}commute\ xs\ {\isasymequiv} {\isasymforall}x\ y{\isachardot}\ {\isacharbraceleft}x{\isacharcomma}\ y{\isacharbraceright}\ {\isasymsubseteq}\ set\ xs\ {\isasymlongrightarrow}\ x\ {\isasymparallel}\ y\ {\isasymlongrightarrow}\ {\isasymlangle}x{\isasymrangle}{\isasymrhd}{\isasymlangle}y{\isasymrangle}\ {\isacharequal}\ {\isasymlangle}y{\isasymrangle}{\isasymrhd}{\isasymlangle}x{\isasymrangle}{\isachardoublequoteclose}
\end{isabellebody}
\vspace{0.35em}
Given this definition, we can now state and prove our main theorem, $\isa{convergence}$.
This theorem states that two hb-consistent lists of distinct operations, which are permutations of each other and in which concurrent operations commute, have the same interpretation:
\vspace{0.35em}
\begin{isabellebody}
\ \ \ \ \ \ \ \ \isacommand{theorem} convergence{\isacharcolon}\isanewline
\ \ \ \ \ \ \ \ \ \ \isakeyword{assumes}\ {\isachardoublequoteopen}set\ xs\ {\isacharequal}\ set\ ys{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}concurrent{\isacharunderscore}ops{\isacharunderscore}commute\ xs{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}concurrent{\isacharunderscore}ops{\isacharunderscore}commute\ ys{\isachardoublequoteclose}\isanewline
\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \isakeyword{and}\ {\isachardoublequoteopen}distinct\ xs{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}distinct\ ys{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}hb{\isacharunderscore}consistent\ xs{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}hb{\isacharunderscore}consistent\ ys{\isachardoublequoteclose}\isanewline
\ \ \ \ \ \ \ \ \ \ \isakeyword{shows}\ {\isachardoublequoteopen}apply{\isacharunderscore}operations\ xs\ {\isacharequal}\ apply{\isacharunderscore}operations\ ys{\isachardoublequoteclose}
\end{isabellebody}
\vspace{0.35em}
\noindent
A fully mechanised proof of this theorem can be found in the supplementary material of this paper.
Although this theorem may seem ``obvious'' at first glance---commutativity allows the operation order to be permuted---it is more subtle than it seems.
The difficulty arises because operations may succeed when applied to some state, but fail when applied to another state (for example, deleting an element that does not exist in the state).
We find it interesting that it is nevertheless sufficient for the definition of $\isa{concurrent-ops-commute}$ to be expressed only in terms of the Kleisli arrow composition, and without explicitly referring to the state.

\subsection{Formalising Strong Eventual Consistency}\label{sect.abstract.sec.spec}

Besides convergence, another required property of SEC is \emph{progress}: if a valid operation was issued on one node, then applying that operation on other nodes must also succeed---that is, the execution must not become stuck in an error state.
Although the type signature of the interpretation function allows operations to fail, we need to prove that in all $\isa{hb-consistent}$ executions such failure never actually occurs.
We capture the combined requirements of convergence and progress in the $\isa{strong-eventual-consistency}$ locale, which extends $\isa{happens-before}$:
\vspace{0.35em}
\begin{isabellebody}
\ \ \ \ \ \ \ \ \isacommand{locale}\ strong{\isacharunderscore}eventual{\isacharunderscore}consistency\ {\isacharequal}\ happens{\isacharunderscore}before\ {\isacharplus}\ \isakeyword{fixes}\ op{\isacharunderscore}history\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharprime}oper\ list\ {\isasymRightarrow}\ bool{\isachardoublequoteclose}\ \isakeyword{and}\ initial{\isacharunderscore}state\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharprime}state{\isachardoublequoteclose}\isanewline
\ \ \ \ \ \ \ \ \ \ \isakeyword{assumes}\ causality{\isacharcolon}\ \ \ \ \ {\isachardoublequoteopen}op{\isacharunderscore}history\ xs\ {\isasymLongrightarrow}\ hb{\isacharunderscore}consistent\ xs{\isachardoublequoteclose}\isanewline
\ \ \ \ \ \ \ \ \ \ \ \ \ \ \isakeyword{and}\ distinctness{\isacharcolon}\ \ {\isachardoublequoteopen}op{\isacharunderscore}history\ xs\ {\isasymLongrightarrow}\ distinct\ xs{\isachardoublequoteclose}\isanewline
\ \ \ \ \ \ \ \ \ \ \ \ \ \ \isakeyword{and}\ trunc{\isacharunderscore}history{\isacharcolon}\ {\isachardoublequoteopen}op{\isacharunderscore}history{\isacharparenleft}xs{\isacharat}{\isacharbrackleft}x{\isacharbrackright}{\isacharparenright}\ {\isasymLongrightarrow}\ op{\isacharunderscore}history\ xs{\isachardoublequoteclose}\isanewline
\ \ \ \ \ \ \ \ \ \ \ \ \ \ \isakeyword{and}\ commutativity{\isacharcolon}\ {\isachardoublequoteopen}op{\isacharunderscore}history\ xs\ {\isasymLongrightarrow}\ concurrent{\isacharunderscore}ops{\isacharunderscore}commute\ xs{\isachardoublequoteclose}\isanewline
\ \ \ \ \ \ \ \ \ \ \ \ \ \ \isakeyword{and}\ no{\isacharunderscore}failure{\isacharcolon}\ \ \ \ {\isachardoublequoteopen}op{\isacharunderscore}history{\isacharparenleft}xs{\isacharat}{\isacharbrackleft}x{\isacharbrackright}{\isacharparenright}\ {\isasymLongrightarrow}\ apply{\isacharunderscore}operations\ xs\ initial{\isacharunderscore}state\ {\isacharequal}\ Some\ state\ {\isasymLongrightarrow}\ {\isasymlangle}x{\isasymrangle}\ state\ {\isasymnoteq}\ None{\isachardoublequoteclose}
\end{isabellebody}
\vspace{0.35em}
Here, $\isa{op-history}$ is an abstract predicate describing any valid operation history of some replication algorithm.
This locale serves as a concise summary of the properties that we require in order to achieve SEC, and from these assumptions and the theorem above we easily obtain the two safety properties of SEC as theorems:
\vspace{0.35em}
\begin{isabellebody}
\ \ \ \ \ \ \ \ \isacommand{theorem}\ {\isacharparenleft}\isakeyword{in}\ strong{\isacharunderscore}eventual{\isacharunderscore}consistency{\isacharparenright}\ sec{\isacharunderscore}convergence{\isacharcolon}\isanewline
\ \ \ \ \ \ \ \ \ \ \isakeyword{assumes}\ {\isachardoublequoteopen}set\ xs\ {\isacharequal}\ set\ ys{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}op{\isacharunderscore}history\ xs{\isachardoublequoteclose}\ \isakeyword{and}\ {\isachardoublequoteopen}op{\isacharunderscore}history\ ys{\isachardoublequoteclose}\isanewline
\ \ \ \ \ \ \ \ \ \ \isakeyword{shows}\ \ \ {\isachardoublequoteopen}apply{\isacharunderscore}operations\ xs\ {\isacharequal}\ apply{\isacharunderscore}operations\ ys{\isachardoublequoteclose}
\end{isabellebody}
\vspace{0.35em}
\begin{isabellebody}
\ \ \ \ \ \ \ \ \isacommand{theorem}\ {\isacharparenleft}\isakeyword{in}\ strong{\isacharunderscore}eventual{\isacharunderscore}consistency{\isacharparenright}\ sec{\isacharunderscore}progress{\isacharcolon}\isanewline
\ \ \ \ \ \ \ \ \ \ \isakeyword{assumes}\ {\isachardoublequoteopen}op{\isacharunderscore}history\ xs{\isachardoublequoteclose}\isanewline
\ \ \ \ \ \ \ \ \ \ \isakeyword{shows}\ \ \ {\isachardoublequoteopen}apply{\isacharunderscore}operations\ xs\ initial{\isacharunderscore}state\ {\isasymnoteq}\ None{\isachardoublequoteclose}
\end{isabellebody}
\vspace{0.35em}
Thus, in order to prove SEC for some replication algorithm, we only need to show that the five assumptions of the $\isa{strong-eventual-consistency}$ locale are satisfied.
As we shall see in Section~\ref{sect.network}, the first three assumptions are satisfied by our network model, and do not require any algorithm-specific proofs.
For individual algorithms we only need to prove the $\isa{commutativity}$ and $\isa{no-failure}$ properties, and we show how to do this in Sections~\ref{sect.rga} and~\ref{sect.simple.crdts}.
